The brightness and spatial distributions of terrestrial radio sources

### 2012-12-17 articles/Article-RFI-distribution/article-rfi-distribution.tex
Implemented Saleems comments

\begin{abstract}
Feeble undetected sources of radio-frequency interference (RFI) might become visible after integration of long radio observations if they are coherently present over time. Thereby, they might occlude the astronomical signal of interest. This issue is especially important for epoch of reionization (EoR) projects that try to detect the faint redshifted HI signals of the earliest objects in the Universe. We explore the low-frequency RFI situation by studying the $\log N$--$\log S$ histograms of data observed with LOFAR. The histograms shows a RFI distribution that follows a power-law distribution with an estimated exponent around -1.6 to -1.5. With several assumptions, this can be explained with a uniform distribution of terrestrial radio sources whose radiation follows existing propagation models. Extrapolation of the power law would imply that the current LOFAR EoR observations should be severely RFI limited if these sources are coherently present. Since these observations have achieved the thermal (sky) noise limit, a logical implication is that most of the RFI sources do not add coherently.
\end{abstract}

@@

\section{Introduction}
Radio astronomy concerns itself with the observation of radiation from celestial sources at radio wavelengths. However, astronomical radio observations can be affected by radio-frequency interference (RFI), which makes it difficult to calibrate the instrument and achieve high sensitivities \citep{impact-of-warc79,interference-and-radioastronomy-1991,interference-model-lemmon,rfi-mitigation-overview-fridman-baan}. Many techniques have been designed to mitigate its effect, such as detection and flagging of data \citep{chi-square-time-blanking-weber, multichannel-rfi-mitigation, exoplanet-detection-with-rfi, wsrt-rfims, pulse-blanking, effelsberg-rfi-mitigation, LOFAR-RFI-pipeline}, adaptive cancellation techniques \citep{adaptive-cancellation,post-correlation-reference-signal} and spatial filtering \citep{multichannel-rfi-mitigation, ellingson-spatial-nulling-2002, hampson-spatial-nulling-2002, boonstra-dissertation, spatial-filtering-parkes-multibeam-for-pulses, post-correlation-filtering}. In most cases, these methods excise enough of the interference to calibrate and image the data.

Regular radio observations record one or a few sessions of typical a few hours and results are integrated. In these cases, it often suffices to only excise interference that is apparent and thus above the noise to reach the thermal noise limit of the instrument. A new challenge arises when one desires much deeper observations, and hundreds of hours of observations need to be integrated. In such a case, feeble interference caused by stationary sources might not manifest itself above the noise in individual observations, but is coherently present in the data. Subsequently, when averaging these data, the interference might become apparent and occlude the signal of interest. This is very relevant for 21-cm Epoch of Reionisation (EoR) experiments, because these involve long integration times. Several such experiments are underway, both to measure the angular power spectrum \citep{gmrt-eor-2011-paciga,de-bruyn-eor-ursi-2011,eor-paper-2011-jacobs,eor-mwa-2012-williams} or global signal \citep{edges}. For these experiments, it is important to know the possible effect of low-level interferences on the data, as these might overshadow or alter the signal of interest. In this article, we will connect new insights about RFI to the angular Epoch of Reionisation experiment that is using the Low-Frequency Array \citep{de-bruyn-eor-ursi-2011}.

@@

In Sect.~\ref{sec:distribution-prediction}, we predict the terrestrial interfering source distribution based on various assumptions. Sect.~\ref{sec:distribution-methods} presents the methods that we use to generate and analyse the histograms. This is followed in Sect.~\ref{sec:dist-data} by a short description of the two LOFAR data sets that have been used to perform the experiment. The results of analysing the sets are presented in Sect.~\ref{sec:dist-results}. Finally, in Sect.~\ref{sec:dist-discussion} the results are discussed and conclusions are drawn.

@@

We assume that the source is observed fully coherent, but a possible decoherence factor can be absorbed in the gains. Due to the small bandwidth of most interfering sources, most RFI will be received coherently, because of the narrow-band condition. For example, if the bandwidth of the signal $\Delta\nu=1$ kHz, the narrow-band condition $\Delta \nu \ll (2\pi \tau)^{-1}$ with correlation delay $\tau$ will hold for baselines up to a few km, because it holds as long as the baseline length is significantly less than $\Delta x = c / (2 \pi \Delta \nu) \approx$ 50 km.

@@

Next, we assume that all sources have equal constant strength $I$. Moreover, we assume that the sources have a uniform spatial distribution in a two-dimensional plane. This is obviously a simplification, as the sources are actually distributed on the surface of a sphere. Beyond some distance, the Earth will also partly block the path between transmitter and receiver. Therefore, the assumption of uniformly distributed sources on a two-dimensional plane is only approximately valid and only for a limited distance. Using these assumptions, we can express the expected cumulative frequency density of sources $f_I(r)$ at distance $r$ as
\begin{equation} \label{eq:cumulative-distance-function}
F_{\textrm{distance} \le r}(r)=c\pi r^2,
\end{equation}
for some constant $c$ that represents the number of sources per unit area. In other words, we will observe $F$ sources that are at most a distance of $r$ away.

@@

The cumulative number of sources $F_{\textrm{amplitude} \le S}$ that have an amplitude of at most $S$ can now be calculated with
\begin{align} \label{eq:cumulative-rfi-distribution}
 F_{\textrm{amplitude} \le S}(S) = & F_{\textrm{distance} \ge r}(\mathcal{S}^{-1}(S))\\
\notag = & F_{\textrm{distance} \ge r}(\pm\sqrt\frac{Ig}{S})\\
\notag = & N-\frac{c\pi Ig}{S}
\end{align}
where $\mathcal{S}^{-1}$ refers to the inverse of $\mathcal{S}(r)$, i.e., the function that returns the distance $r$ for a given amplitude $S$. Finally, the differential frequency density can be calculated by taking the derivative,
\begin{align} \label{eq:two-dimensional-distribution}
 f_S(S) = & \frac{dF_{\textrm{amplitude} \le S}}{dS} = \frac{c\pi Ig}{S^2}.
\end{align}

@@

A commonly used propagation model is the empirical model determined by \citet{okumura-propagation-model}, which was further developed by \citet{hata-propagation-loss}. \citeauthor{hata-propagation-loss} gives the following analytical estimate for $L_p$, the electromagnetic propagation loss over land:
\begin{align} 
\notag                L_p & = & 69.55 + 26.16 \log_{10} f_c - 13.82 \log_{10} h_b - \\
\label{eq:hata-model} & &  a(h_m) + (44.9 - 6.55 \log_{10} h_b) \log_{10} r, 
\end{align}
where $L_p$ the loss in dB; $f_c$ the radiation frequency in MHz; $h_b$ the height of the transmitting antenna in m; $h_m$ the height of the receiving antenna in m; $r$ the distance between the antennae in m; and $a(h_m)$ a correction factor in dB that corrects for the height of the receiving antenna and the urban density. \citeauthor{hata-propagation-loss} found this model to be representative for frequencies $f_c \sim$~150--1500~MHz, with transmitter heights $h_b \sim$~30--200~m, receiver heights $h_m \sim$~1--10~m and over distances $r\sim$~1--20~km.

Converting from a subtrahend in decibels to a flux density factor $L_S$, and collecting the terms of Eq.~\eqref{eq:hata-model} that are not depending on $r$ in a single variable $\zeta$, results in
\begin{equation} \label{eq:propagation-loss-factor}
 L_S = \zeta r^{4.49 - 0.655 \log_{10} h_b},
\end{equation}
with
\begin{equation}
 \zeta = \frac{f_c^{2.616}}{h_b^{1.382}} - 10^{6.955-\frac{1}{10} a(h_m)}.
\end{equation}
The model is based on urban areas, but corrections are given by \citeauthor{hata-propagation-loss} for sub-urban areas with a lower population density and for open areas. These corrections are independent of $r$, thus would only change $\zeta$. Note that according to Hata's model, the exponent of the power law in Eq.~\eqref{eq:propagation-loss-factor} depends only on the height of the transmitting antenna, i.e., it is independent of frequency, receiver height and urban density. Now, if in Eqs.~\eqref{eq:cumulative-rfi-distribution} and \eqref{eq:two-dimensional-distribution} one replaces the definition of $S(r)$ from Eq.~\eqref{eq:amplitude-fall-off-in-free-space} with one that includes the propagation effects,
\begin{equation}
 S(r) = \frac{\zeta Ig}{r^\eta},
\end{equation}
with $\eta = 0.655 \log_{10} h_b - 4.49$, one finds the frequency density function $f_p$ that considers propagation effects,
% S(r):=I*g/(r^alpha);
% assume(g>0,I>0,S>0);
% diff(c*%pi*r^2,S),solve(S(r)=S,r);
\begin{align}
f_p(S)
= & \frac{d}{dS} c \pi \left( \frac{\zeta Ig }{S} \right)^{2/\eta}
= \frac{c 2\pi}{\eta S} \left( \frac{\zeta Ig}{S} \right)^{2/\eta}.
\end{align}
Consequently, due to non-free space propagation effects, the observed log-log histogram is predicted to have a $\frac{2}{\eta}-1$ slope. By substituting $\eta$, one finds
\begin{equation}
\textrm{slope}(h_b) = \frac{1}{0.3275 \log_{10} h_b - 2.245} - 1.
\end{equation}
\begin{figure}
\begin{center}\hspace{-1cm}\includegraphics[width=6cm]{img/transmitter-height/plot-transmitter-height-trimmed}
\caption{Effect of transmitter height on the slope of a log-log histogram.}
\label{fig:plot-transmitter-height}
\end{center}
\end{figure}
This is valid for transmitters that have a height of 30--200 m, the range over which Hata's model was defined. This yields estimated distribution slopes of $-1.57$ and $-1.67$ for $30$ m and $200$ m high transmitters respectively. In Figure~\ref{fig:plot-transmitter-height}, the slope value is plotted as a function of the transmitter height, including extrapolated values for transmitter heights down to 1 m.

@@

Fig.~\ref{fig:rayleigh-and-rfi-combined} shows that the left side of the graph corresponds with the original Rayleigh distribution. From the right side of the histogram, we are able to estimate $Ig/r_L^2$, as the placing of the power-law curve is independent of the Rayleigh distribution for large $S$. Methods to constrain the parameters of the RFI distribution will be discussed in \S\ref{sec:rfi-distribution-constraints}. The curves in Fig.~\ref{fig:rayleigh-and-rfi-combined} are from histograms of samples that are all contaminated by both noise and RFI. In practice, we can not make this distinction 100\% accurately, as RFI detectors have a limited accuracy. Consequently, the observed histogram will be the sum of two types of histograms: the first histogram being contaminated by both noise and RFI, the second one only by noise.

@@

Another effect that is absorbed in $g$, is the beam of the instrument. At the point of writing, LOFAR beam models are still being developed and are not yet well parametrized. However, since it is likely that most RFI sources are observed at the edges of the beam, it can be expected that the beam will have a benign effect on the histogram of an observation, comparable with the effect of the frequency response.

@@

Assume also that the histogram contains $N_\textrm{part}$ (RFI) samples with amplitude $> S_1$, as sketched in Fig.~\ref{fig:explanation-constraints}. The hypothetical upper limit $S_U$ of the distribution can now be found, i.e., the highest amplitude that would be observed when the distribution follows the power law up to that point, by solving
\begin{equation} \label{eq:upper-limit-constrained}
 \int\limits_{S_1}^{S_U} g(S) dS = N_\textrm{part}.
\end{equation}
In practice, the observed histogram will break down beyond some amplitude because of several reasons. First and most importantly, the samples itself are digitized with an analogue-to-digital converter (ADC) with limited range. Second, we observe for a limited time and the frequency count is discrete. Because the chance of finding a sample with a very high intensity is low, it is unlikely to observe samples beyond some amplitude within the finite observing time. Finally, under the assumption of a uniform spatial distribution of RFI transmitters, samples with very high amplitude would have to be produced by transmitters that are very close to the telescope. However, it is likely that the uniform spatial distribution of transmitters will break down at closer distances.

Solving Eq.~\eqref{eq:upper-limit-constrained} results in
\begin{equation} \label{eq:upper-limit}
S_U = \sqrt[\alpha+1]{\frac{\alpha+1}{\beta} N_\textrm{part} + S_1^{\alpha+1}}.
\end{equation}
In some cases there will be no solution for $S_U$. The reason for this is that the integral of the distribution function converges, and the total number of samples is finite even when $S \rightarrow \infty$. If $N_\textrm{part}$ exceeds the integrated value of the fitted distribution function, there is no solution. This can happen when the empirical distribution is not limited at the high end, or when it contains features that are not in the model.

@@

\subsection{Error analysis}
An estimate for the standard deviation of the slope estimator $\hat \alpha$ can be found by calculating $\textrm{SE}(\hat \alpha)$, the \emph{standard error} of $\hat \alpha$. The standard error of the slope of a straight line \citep[pp. 32--35]{acton-analaysis-of-straight-lines} is given by
\begin{equation} \label{eq:stderr-slope}
 \textrm{SE}(\hat \alpha) = \sqrt{\frac{SS_{yy}-\hat\alpha SS_{xy}}{\left(n - 2\right) SS_{xx}}},
\end{equation}
where $SS_{xx}$, $SS_{xy}$ and $SS_{yy}$ are the sums of squares, e.g., $SS_{xy}=\sum_{i=1}^n (x_i - \bar x) (y_i - \bar y)$ and $n$ is the number of samples. However, we found that this is not a representable error in our case, because the errors in the slope are not normally distributed. The noise in the part of the histogram over which the slope is calculated, is very small due to the large amount of samples. Consequently, the estimated standard deviation of $\hat \alpha$ is very low. Nevertheless, when the slope is calculated over subsections of the original range over which the slope is calculated, one finds the line is not completely straight and the slope is changing more than is predicted by the standard error. Therefore, we introduce an error estimate $\epsilon_\alpha$ that quantifies a normalized standard deviation of the slope over the range. This error is formed by calculating the slope over $n_\alpha$ smaller subranges in the histogram, creating $n_\alpha$ estimates $\alpha_i$. If the errors in $\alpha_i$ are normally distributed with mean zero, the standard deviation over the full range will be $\sqrt{n_\alpha}$ times smaller. Therefore, an estimate of the variance of $\hat \alpha$ can be calculated with
\begin{equation}
 \epsilon_{\hat\alpha} = \sqrt{\frac{\sum \left( \alpha_i - \hat\alpha \right)^2 }{n^2_\alpha - n_\alpha}}.
\end{equation}
This estimate is slightly depending on the number of subranges that is used, $n_\alpha$, because the errors are not Gaussian distributed, but we found that $\epsilon_{\hat\alpha}$ is more representative than the standard error of $\hat\alpha$.

The Hill estimator of Eq.~\eqref{eq:hill} is a different estimation method for the exponent in the power-law distribution, and yields therefore also a different standard error. The standard error of the Hill estimator is \citep{power-law-distribution}
\begin{equation} \label{eq:stderr-hill}
 \textrm{SE}(\hat \alpha_H) = \frac{-\alpha - 1}{\sqrt{n}} + \mathcal{O}(\frac{1}{n}).
\end{equation}
Because the number of samples is very large ($>10^{11}$), the $\mathcal{O}$-term will be very small. Therefore, we will calculate the quantity without the term. As with the standard error for the slope of a straight line in Eq.~\eqref{eq:stderr-slope}, the standard error for the Hill estimator yields very small quantities. Again, this is because it assumes the underlying power law has a fixed exponential, while in our case the power law seems to vary. Therefore, this value is given only for completeness.

@@

If we go back to Fig.~\ref{fig:histogram-passband-corrected}, we see that in the LBA the power law is stable for about three orders of magnitude, and one order more in the HBA. Fig.~\ref{fig:plot-slopes} shows the slope of the log-log plot as a function of amplitude, which was constructed by performing linear regression in a sliding window, with a window size of 1 decade. The HBA shows very little structure in the slope, but the HBA is less stable and shows some features in its power-law part. Linear regression on the power-law part of the log-log plot results in a slope of $-1.62$ for the LBA and $-1.53$ for the HBA. These and the other derived quantities have been summarized in Table~\ref{table:dist-data-quantities}. Although the HBA slope does not show any other significant features besides the Rayleigh and power-law curves, the LBA power law ends with a bulge around an amplitude of $10^6$. This bulge is caused by a very strong source affecting lots of samples, and is a single outlier in the spatial distribution. We found this is caused by RFI observed for about an hour in the late afternoon in the lower LBA frequency regime, around $30$--$40$~MHz. Leaving this frequency range out flattens the bulge significantly, but does not completely eliminate it, because the source turned the receivers in non-linear state, causing leakage at lower intensity levels in the other sub-bands. Unlike linear regression, the fitting region of the Hill estimator is not limited at the high end. Consequently, because of the bulge, the Hill estimator evaluates for the LBA into a slope that is less steep, with a value of $-1.53$. For the HBA set, the Hill estimator is equal to the $-1.53$ value found by linear regression.

@@

On the assumption that the histogram is zero below amplitude $S_L$, we find that $S_L=21$ mJy for the LBA and $S_L=6.2$ mJy for the HBA (see Table~\ref{table:dist-data-quantities}). If instead it is assumed that the histogram has a uniform distribution below some amplitude $\tilde{S_L}$, we find that the amplitude at which the power-law distribution breaks down is approximately a factor two higher. The two different assumptions on how the power-law distribution breaks down have a small effect on $E(S_\textrm{leak})$, the expected value of the leaked RFI. By using $\tilde{S}_L$ instead of $S_L$, it is a few percent lower. By assuming a 100\% RFI occupancy, we find that the expected value of leaked RFI is $484$--$496$ mJy for the LBA and $167$--$171$ mJy. By assuming $10$\% occupancy, the value for $E(S_\textrm{leak})$ is about $25$\% reduced. The RFI occupancy only starts to have a significant effect on $E(S_\textrm{leak})$ if it is well below $10$\%.

@@

For the EoR project with $1$~MHz resolution, Eq.~\ref{eq:rfi-noise-level} results in $\sigma_\textrm{RFI}\approx 4.7$ mJy. However, the first EoR results of observations of one day have approximately reached the thermal noise of about $1.7$ mJy per 0.2~MHz subband (Yatawatta et al., \textit{in preparation}), and the resulting images show no signs of RFI. This implies that either the upper limit is far from the actual RFI situation, or Eq.~\ref{eq:rfi-noise-level} is not applicable to most of the RFI that is observed with LOFAR. In the following section we will discuss effects that could cause a reduced contribution of RFI.

>>>>

\begin{abstract}
Feeble undetected sources of radio-frequency interference (RFI) might become visible after integration of long radio observations if they are coherently present over time. Thereby, they might obstruct the detection of the astronomical signal of interest. This issue is especially important for Epoch of Reionization (EoR) projects that try to detect the faint redshifted HI signals of the earliest objects in the Universe. We explore the low-frequency RFI situation by studying the $\log N$--$\log S$ histograms of data observed with LOFAR. The histograms shows an RFI distribution that follows a power-law distribution with an estimated exponent around -1.6 to -1.5. With several assumptions, this can be explained with a uniform distribution of terrestrial radio sources whose radiation follows existing propagation models. Extrapolation of the power law would imply that the current LOFAR EoR observations should be severely RFI limited if these sources are coherently present. Since these observations have achieved the thermal (sky) noise limit, a logical implication is that most of the RFI sources do not add coherently.
\end{abstract}

@@

\section{Introduction}
Radio astronomy concerns itself with the observation of radiation from celestial sources at radio wavelengths. However, astronomical radio observations can be affected by radio-frequency interference (RFI), which makes it difficult to calibrate the instrument and achieve high sensitivities \citep{impact-of-warc79,interference-and-radioastronomy-1991,interference-model-lemmon,rfi-mitigation-overview-fridman-baan}. Many techniques have been designed to mitigate this effect, such as detection and flagging of data \citep{chi-square-time-blanking-weber, multichannel-rfi-mitigation, exoplanet-detection-with-rfi, wsrt-rfims, pulse-blanking, effelsberg-rfi-mitigation, LOFAR-RFI-pipeline}, adaptive cancellation techniques \citep{adaptive-cancellation,post-correlation-reference-signal} and spatial filtering \citep{multichannel-rfi-mitigation, ellingson-spatial-nulling-2002, hampson-spatial-nulling-2002, boonstra-dissertation, spatial-filtering-parkes-multibeam-for-pulses, post-correlation-filtering}. In most cases, these methods excise enough of the interference to calibrate and image the data.

Typical radio observations record a few hours of data, and the results are integrated. In these cases, it often suffices to only excise interference that is apparent and thus above the noise to reach the thermal noise limit of the instrument. A new challenge arises, however, when one desires much deeper observations, and hundreds of hours of observations need to be integrated. In such a case, feeble interference caused by stationary sources might not manifest itself above the noise in individual observations, but is coherently present in the data. Subsequently, when averaging these data, the interference might become apparent and occlude the signal of interest. This is very relevant for the 21-cm Epoch of Reionisation (EoR) experiments, because they involve long integration times. Several such experiments are underway, to either measure the angular power spectrum \citep{gmrt-eor-2011-paciga,de-bruyn-eor-ursi-2011,eor-paper-2011-jacobs,eor-mwa-2012-williams} or the global signal \citep{edges}. For these experiments, it is important to know the possible effect of low-level interferences on the data, as these might overshadow or alter the signal of interest. In this article, we will connect new insights about RFI to the angular Epoch of Reionisation experiment that is using the Low-Frequency Array \citep{de-bruyn-eor-ursi-2011}.

@@

This paper is organised as follows: in Sect.~\ref{sec:distribution-prediction}, we predict the terrestrial interfering source distribution based on various assumptions. Sect.~\ref{sec:distribution-methods} presents the methods that we use to generate and analyse the histograms. This is followed in Sect.~\ref{sec:dist-data} by a short description of the two LOFAR data sets that have been used to perform the experiment. The results of analysing the sets are presented in Sect.~\ref{sec:dist-results}. Finally, in Sect.~\ref{sec:dist-discussion} the results are discussed and conclusions are drawn.

@@

We assume that the source observed is fully coherent, but a possible decoherence factor can be absorbed in the gains. Due to the small bandwidth of most interfering sources, most RFI will be received coherently, because of the narrow-band condition. For example, if the bandwidth of the signal $\Delta\nu=1$ kHz, the narrow-band condition $\Delta \nu \ll (2\pi \tau)^{-1}$ with correlation delay $\tau$ will hold for baselines up to a few km, because it holds as long as the baseline length is significantly less than $\Delta x = c / (2 \pi \Delta \nu) \approx$ 50 km.

@@

Next, we assume that all sources have equal constant strength $I$ and follow a uniform spatial distribution in a two-dimensional plane. This is obviously a simplification, as the sources are actually distributed on the surface of a sphere. Hence, beyond some distance, the Earth will partly block the path between transmitter and receiver. Therefore, the assumption of uniformly distributed sources on a two-dimensional plane is only approximately valid. Using these assumptions, we can express the expected cumulative frequency density of sources $f_I(r)$ at distance $r$ as
\begin{equation} \label{eq:cumulative-distance-function}
F_{\textrm{distance} \le r}(r)=c\pi r^2,
\end{equation}
for some constant $c$ that represents the number of sources per unit area. In other words, we will observe $F$ sources that are at most a distance of $r$ away.

@@

The cumulative number of sources $F_{\textrm{amplitude} \le S}$ that have an amplitude of at most $S$ can now be calculated with
\begin{align} \label{eq:cumulative-rfi-distribution}
 F_{\textrm{amplitude} \le S}(S) = & F_{\textrm{distance} \ge r}(\mathcal{S}^{-1}(r))\\
\notag = & F_{\textrm{distance} \ge r}(\pm\sqrt\frac{Ig}{S})\\
\notag = & N-\frac{c\pi Ig}{S}
\end{align}
where $\mathcal{S}^{-1}$ is the inverse of $\mathcal{S}(r)$, i.e., the function that returns the distance $r$ for a given amplitude $S$. Finally, the differential frequency density can be calculated by taking the derivative,
\begin{align} \label{eq:two-dimensional-distribution}
 f_S(S) = & \frac{dF_{\textrm{amplitude} \le S}}{dS} = \frac{c\pi Ig}{S^2}.
\end{align}

@@

A commonly used propagation model is the empirical model determined by \citet{okumura-propagation-model}, which was further developed by \citet{hata-propagation-loss}. The original model is based on urban areas, but corrections are given by \citeauthor{hata-propagation-loss} for sub-urban areas with a lower population density and for open areas. \citeauthor{hata-propagation-loss} gives the following analytical estimate for $L_p$, the electromagnetic propagation loss over land:
\begin{align} 
\notag                L_p & = & 69.55 + 26.16 \log_{10} f_c - 13.82 \log_{10} h_b - \\
\label{eq:hata-model} & &  a(h_m) + (44.9 - 6.55 \log_{10} h_b) \log_{10} r, 
\end{align}
where $L_p$ the loss in dB; $f_c$ the radiation frequency in MHz; $h_b$ the height of the transmitting antenna in meters; $h_m$ the height of the receiving antenna in meters; $r$ the distance between the antennae in meters; and $a(h_m)$ a correction factor in dB that corrects for the height of the receiving antenna and the urban density. \citeauthor{hata-propagation-loss} found this model to be representative for frequencies $f_c \sim$~150--1500~MHz, with transmitter heights $h_b \sim$~30--200~m, receiver heights $h_m \sim$~1--10~m and over distances $r\sim$~1--20~km.

Converting from a subtrahend in decibels to a flux density factor $L_S$, and collecting the terms of Eq.~\eqref{eq:hata-model} that are not depending on $r$ in a single variable $\zeta$, results in
\begin{equation} \label{eq:propagation-loss-factor}
 L_S = \zeta r^{4.49 - 0.655 \log_{10} h_b},
\end{equation}
with
\begin{equation}
 \zeta = \frac{f_c^{2.616}}{h_b^{1.382}} - 10^{6.955-\frac{1}{10} a(h_m)}.
\end{equation}
Note that according to Hata's model, the exponent of the power law in Eq.~\eqref{eq:propagation-loss-factor} depends only on the height of the transmitting antenna, i.e., it is independent of frequency, receiver height and urban density. Now, if in Eqs.~\eqref{eq:cumulative-rfi-distribution} and \eqref{eq:two-dimensional-distribution} one replaces the definition of $S(r)$ from Eq.~\eqref{eq:amplitude-fall-off-in-free-space} with one that includes the propagation effects,
\begin{equation}
 S(r) = \frac{\zeta Ig}{r^\eta},
\end{equation}
with $\eta = 0.655 \log_{10} h_b - 4.49$, one finds the frequency density function $f_p$ that considers propagation effects,
% S(r):=I*g/(r^alpha);
% assume(g>0,I>0,S>0);
% diff(c*%pi*r^2,S),solve(S(r)=S,r);
\begin{align}
f_p(S)
= & \frac{d}{dS} c \pi \left( \frac{\zeta Ig }{S} \right)^{2/\eta}
= \frac{c 2\pi}{\eta S} \left( \frac{\zeta Ig}{S} \right)^{2/\eta}.
\end{align}
Consequently, due to non-free space propagation effects, the observed log-log histogram is predicted to have a $\frac{2}{\eta}-1$ slope. By substituting $\eta$, one finds
\begin{equation}
\textrm{slope}(h_b) = \frac{1}{0.3275 \log_{10} h_b - 2.245} - 1.
\end{equation}
\begin{figure}
\begin{center}\hspace{-1cm}\includegraphics[width=6cm]{img/transmitter-height/plot-transmitter-height-trimmed}
\caption{Effect of transmitter height on the slope of a log-log histogram.}
\label{fig:plot-transmitter-height}
\end{center}
\end{figure}
This is valid for transmitters that have a height of 30--200 m, the range over which Hata's model was defined. This yields estimated distribution slopes of $-1.57$ and $-1.67$ for $30$ m and $200$ m high transmitters respectively. In Figure~\ref{fig:plot-transmitter-height}, the slope value is plotted as a function of the transmitter height, including extrapolated values for transmitter heights down to 1 m.

@@

Fig.~\ref{fig:rayleigh-and-rfi-combined} shows that the left side of the graph corresponds with the original Rayleigh distribution. From the right side of the histogram, we are able to estimate $Ig/r_L^2$, as the placing of the power-law curve is independent of the Rayleigh distribution for large $S$. Methods to constrain the parameters of the RFI distribution will be discussed in \S\ref{sec:rfi-distribution-constraints}. The curves in Fig.~\ref{fig:rayleigh-and-rfi-combined} are from histograms of samples that are all contaminated by both noise and RFI. In practice, we can not make this distinction accurately, as RFI detectors have a limited accuracy. Consequently, the observed histogram will be the sum of two types of histograms: the first histogram being contaminated by both noise and RFI, the second one only by noise.

@@

Another effect that is absorbed in $g$, is the beam of the instrument. At the point of writing, LOFAR beam models are still being developed and are not yet well parametrized. However, since it is likely that most RFI sources are observed at the edges of the beam, it can be expected that the beam will have a limited effect on the histogram of an observation, comparable with the effect of the frequency response.

@@

Assume also that the histogram contains $N_\textrm{part}$ (RFI) samples with amplitude $> S_1$, as sketched in Fig.~\ref{fig:explanation-constraints}. The hypothetical upper limit $S_U$ of the distribution can now be found, i.e., the highest amplitude that would be observed when the distribution follows the power law up to that point, by solving
\begin{equation} \label{eq:upper-limit-constrained}
 \int\limits_{S_1}^{S_U} g(S) dS = N_\textrm{part}.
\end{equation}
In practice, the observed histogram will break down beyond some amplitude because of several reasons. First and most importantly, the samples themselves are digitized with an analogue-to-digital converter (ADC) with limited range. Second, we observe for a limited time and the frequency count is discrete. Because the chance of finding a sample with a very high intensity is low, it is unlikely to observe samples beyond some amplitude within the finite observing time. Finally, under the assumption of a uniform spatial distribution of RFI transmitters, samples with very high amplitude would have to be produced by transmitters that are very close to the telescope. However, it is likely that the uniform spatial distribution of transmitters will break down at closer distances.

Solving Eq.~\eqref{eq:upper-limit-constrained} results in
\begin{equation} \label{eq:upper-limit}
S_U = \sqrt[\alpha+1]{\frac{\alpha+1}{\beta} N_\textrm{part} + S_1^{\alpha+1}}.
\end{equation}
In some cases there will be no solution for $S_U$. This situation arises because the integral of the distribution function is finite even when $S_U \rightarrow \infty$. If $N_\textrm{part}$ exceeds the integrated value of the fitted distribution function, there is no solution. This can happen when the empirical distribution is not limited at the high end, or when it contains features that are not in the model.

@@

\subsection{Error analysis}
An estimate for the standard deviation of the slope estimator $\hat \alpha$ can be found by calculating $\textrm{SE}(\hat \alpha)$, the \emph{standard error} of $\hat \alpha$. The standard error of the slope of a straight line \citep[pp. 32--35]{acton-analaysis-of-straight-lines} is given by
\begin{equation} \label{eq:stderr-slope}
 \textrm{SE}(\hat \alpha) = \sqrt{\frac{SS_{yy}-\hat\alpha SS_{xy}}{\left(n - 2\right) SS_{xx}}},
\end{equation}
where $SS_{xx}$, $SS_{xy}$ and $SS_{yy}$ are the sums of squares, e.g., $SS_{xy}=\sum_{i=1}^n (x_i - \bar x) (y_i - \bar y)$ and $n$ is the number of samples. However, we found that this is not a representative error in our case, because the errors in the slope are not normally distributed. The noise in the part of the histogram over which the slope is calculated, is very small due to the large amount of samples. Consequently, the estimated standard deviation of $\hat \alpha$ is very low. Nevertheless, when the slope is calculated over subsections of the original range over which the slope is calculated, one finds the line is not completely straight and the slope is changing more than is predicted by the standard error. Therefore, we introduce an error estimate $\epsilon_\alpha$ that quantifies a normalized standard deviation of the slope over the range. This error is formed by calculating the slope over $n_\alpha$ smaller subranges in the histogram, creating $n_\alpha$ estimates $\alpha_i$. If the errors in $\alpha_i$ are normally distributed with mean zero, the standard deviation over the full range will be $\sqrt{n_\alpha}$ times smaller. Therefore, an estimate of the variance of $\hat \alpha$ can be calculated with
\begin{equation}
 \epsilon_{\hat\alpha} = \sqrt{\frac{\sum \left( \alpha_i - \hat\alpha \right)^2 }{n^2_\alpha - n_\alpha}}.
\end{equation}
This estimate is slightly depending on the number of subranges that is used, $n_\alpha$, because the errors are not Gaussian distributed, but we found that $\epsilon_{\hat\alpha}$ is more representative than the standard error of $\hat\alpha$.

The Hill estimator of Eq.~\eqref{eq:hill} is a different estimation method for the exponent in the power-law distribution, and yields therefore also a different standard error. The standard error of the Hill estimator is \citep{power-law-distribution}
\begin{equation} \label{eq:stderr-hill}
 \textrm{SE}(\hat \alpha_H) = \frac{-\alpha - 1}{\sqrt{n}} + \mathcal{O}(\frac{1}{n}).
\end{equation}
Because the number of samples is very large ($>10^{11}$), the $\mathcal{O}$-term will be very small. Therefore, we will calculate the quantity without this term. As with the standard error for the slope of a straight line in Eq.~\eqref{eq:stderr-slope}, the standard error for the Hill estimator yields very small quantities. Again, this is because it assumes the underlying power law has a fixed exponential, while in our case the power law seems to vary. Therefore, this value is given only for completeness.

@@

If we go back to Fig.~\ref{fig:histogram-passband-corrected}, we see that in the LBA the power law is stable for about three orders of magnitude, and one order more in the HBA. Fig.~\ref{fig:plot-slopes} shows the slope of the log-log plot as a function of amplitude, which was constructed by performing linear regression in a sliding window, with a window size of 1 decade. The HBA shows very little structure in the slope, but the LBA is less stable and shows some features in its power-law part. Linear regression on the power-law part of the log-log plot results in a slope of $-1.62$ for the LBA and $-1.53$ for the HBA. These and the other derived quantities have been summarized in Table~\ref{table:dist-data-quantities}. Although the HBA slope does not show any other significant features besides the Rayleigh and power-law curves, the LBA power law ends with a bulge around an amplitude of $10^6$. This bulge is caused by a very strong source affecting lots of samples, and is a single outlier in the spatial distribution. We found this is caused by RFI observed for about an hour in the late afternoon in the lower LBA frequency regime, around $30$--$40$~MHz. Leaving this frequency range out flattens the bulge significantly, but does not completely eliminate it, because the source turned the receivers in non-linear state, causing leakage at lower intensity levels in the other sub-bands. Unlike linear regression, the fitting region of the Hill estimator is not limited at the high end. Consequently, because of the bulge, the Hill estimator evaluates for the LBA into a slope that is less steep, with a value of $-1.53$. For the HBA set, the Hill estimator is equal to the $-1.53$ value found by linear regression.

@@

On the assumption that the histogram is zero below amplitude $S_L$, we find that $S_L=21$ mJy for the LBA and $S_L=6.2$ mJy for the HBA (see Table~\ref{table:dist-data-quantities}). If instead it is assumed that the histogram has a uniform distribution below some amplitude $\tilde{S_L}$, we find that the amplitude at which the power-law distribution breaks down is approximately a factor two higher. The two different assumptions on how the power-law distribution breaks down have a small effect on $E(S_\textrm{leak})$, the expected value of the leaked RFI. By using $\tilde{S}_L$ instead of $S_L$, it is a few percent lower. By assuming a 100\% RFI occupancy, we find that the expected value of leaked RFI is $484$--$496$ mJy for the LBA and $167$--$171$ mJy for the HBA. By assuming $10$\% occupancy, the value for $E(S_\textrm{leak})$ is about $25$\% reduced. The RFI occupancy only starts to have a significant effect on $E(S_\textrm{leak})$ if it is well below $10$\%.

@@

For the EoR project with $1$~MHz resolution, Eq.~\ref{eq:rfi-noise-level} results in $\sigma_\textrm{RFI}\approx 4.7$ mJy. However, the first EoR results of observations of one day have approximately reached the thermal noise of about $1.7$ mJy per 0.2~MHz subband (Yatawatta et al., \textit{submitted}), and the resulting images show no signs of RFI. This implies that either the upper limit is far from the actual RFI situation, or Eq.~\ref{eq:rfi-noise-level} is not applicable to most of the RFI that is observed with LOFAR. In the following section we will discuss effects that could cause a reduced contribution of RFI.
